Hydrodynamics of a Semipermeable Inextensible Membrane Under Flow and Confinement

### 2021-02-14 vesicle_code/docs/permeableVes/semipermeable_vesicle.tex
rewrote time stepping and few changes

To answer these
questions, we focus on the balance between hydrodynamics stresses,
tension, and elastic stress of a semipermeable membrane under various
flowing conditions and confinement (Figure~\ref{fig:sketch}).
%
Using numerical simulations described in \S~\ref{sec:formulation} and
\S~\ref{sec:numerics} we can quantify the effects of water permeability
on the hydrodynamics of a vesicle under mechanical stresses.  Without an
osmotic gradient, we first show that a freely suspended semipermeable
vesicle behaves differently from an impermeable vesicle over a long time
in \S~\ref{sec:free_space}. Next in \S~\ref{sec:confinement} we show
that
%, at normal physiological conditions,
the permeating water flow can be amplified by extreme confinement to give rise to observable change in both  vesicle hydrodynamics and vesicle's water content. 
These results provide insight into quantifying the effects of membrane permeability on vesicle hydrodynamics in microfluidics.
%
%shed light on the dynamic aging process of
%living cells that constantly go through flows and constrictions over a
%period of days.

@@

\subsection{Discretization in Time}
To develop a second-order, non-stiff, adaptive time stepping method, we
first introduce the notation
%\begin{alignat}{3}
%  &\BB\ff = -\frac{d^4}{ds^4} \ff,  \qquad
%  &&\TT\sigma = (\sigma \xx_s)_s, \\
%  &\DD\ff = \xx_s \cdot \ff_s, 
%  &&P\ff = (\ff \cdot \nn) \nn,
%\end{alignat}
\begin{align}
  \BB\ff = -\frac{d^4}{ds^4} \ff,  \qquad
  \TT\sigma = (\sigma \xx_s)_s, \quad
  \DD\ff = \xx_s \cdot \ff_s, \quad
  P\ff = (\ff \cdot \nn) \nn,
  \label{eqn:operators}
\end{align}
for the differential and projection operators. Then,
Equation~\eqref{eqn:vesVelocity} can be written as
\begin{align}
  \dot{\xx} &= \uu_\infty(\xx) + \beta P(\BB\xx + \TT\sigma)
  + \SS (\BB\xx + \TT\sigma), \quad \DD \dot{\xx} = 0.
\end{align}
We note that the operators defined in Equation~\eqref{eqn:operators}, as
well as the single-layer potential defined in Equation~\eqref{eqn:SLP}
all depend on the vesicle shape. A time stepping method that treats
these dependencies implicitly would result in a non-linear set of
equations that would be computationally expensive to solve. Instead, we
use a time stepping method that defines the operators at time $t^N$ and
discretizes the stiff terms due to high-order derivatives
implicitly~\cite{vee-gue-zor-bir2009, qua-bir2014}. In particular, we
use the first-order semi-implicit method
%We note that the integral and differential operators, which all depend
%on the vesicle shape, are non-linear with respect to the vesicle shape.
%We construct a time stepping scheme by letting $\BB^N$ be the bending
%operator due to the vesicle configuration at time $t^N$, and use
%similar notation for the other operators. Then, a first-order
%semi-implicit time stepping method of Equation~\eqref{eqn:vesVelocity}
%is
\begin{alignat}{3}  
  \frac{\xx^{N+1} - \xx^N}{\triangle t} &= \uu_\infty(\xx^N) 
  + \beta P^N\left(\BB^N\xx^{N+1} + \TT^N\sigma^{N+1}\right) 
  + \SS^N\left(\BB^N\xx^{N+1} + \TT^N\sigma^{N+1}\right),  \\
  \DD^N\xx^{N+1} &= 1,
\end{alignat}
and this linear system is solved with matrix-free GMRES.

To achieve second-order accuracy over long time horizons, we modify an
adaptive spectral deferred correction time stepping
method~\cite{qua-bir2016}. Since the area of the vesicle is not
conserved, we only use the length of the vesicle to estimate the error
which determines if a time step is accepted or not and determines the
subsequent time step size.

@@

\section{Hydrodynamics of A semipermeable inextensible membrane under strong confinement\label{sec:confinement}}
Here we consider a semipermeable inextensible membrane under strong confinement: a
long closely-fit channel and a contracting channel. In a narrow
microfluidic channel, a large pressure jump is often used to push the
vesicle through a closely-fit channel~\cite{abk-fai-sto2006}. Our
simulations show that such a configuration can lead to an amplification
of water permeation at a fast time scale of the order of mini seconds. In
addition, we find that a large value of $\beta$ is needed to keep the
membrane tension below the threshold value for membrane poration. Such
high permeability for water may be regarded as an indication that the
lipid bilayer membrane is porated under strong confinement, and our
model inextensible membrane is for a porated vesicle within the Helfrich free energy
framework. As in the previous examples, we fix the membrane length and
examine the enclosed water content as a dynamic consequence of water
flow, membrane shape deformation, and membrane tension distribution.

>>>>

To answer these
questions, we focus on the balance between hydrodynamics stresses,
tension, and elastic stress of a semipermeable membrane under various
flowing conditions and confinement (Figure~\ref{fig:sketch}).
%
Using numerical simulations described in \S~\ref{sec:formulation} and
\S~\ref{sec:numerics} we quantify the effects of water permeability on
the hydrodynamics of a vesicle under mechanical stresses.  Without an
osmotic gradient, we first show that a freely suspended semipermeable
vesicle behaves differently from an impermeable vesicle over a long time
in \S~\ref{sec:free_space}. Next in \S~\ref{sec:confinement} we show
that
%, at normal physiological conditions,
the permeating water flow can be amplified by extreme confinement to give rise to observable change in both  vesicle hydrodynamics and vesicle's water content. 
These results provide insight into quantifying the effects of membrane permeability on vesicle hydrodynamics in microfluidics.
%
%shed light on the dynamic aging process of
%living cells that constantly go through flows and constrictions over a
%period of days.

@@

\subsection{Discretization in Time}
We apply a second-order spectral deferred correction (SDC) method that was
thoroughly tested in previous work~\cite{qua-bir2016}. To control error
over long time horizons, we use the error in length to adaptively adjust
the time step size. SDC requires two steps of a first-order time
stepping method. To allow for large time step sizes, terms involving
high-order derivatives are discretized semi-implicitly. We start by
defining
\begin{align}
  \BB\ff = -\frac{d^4}{ds^4} \ff,  \qquad
  \TT\sigma = (\sigma \xx_s)_s, \quad
  \DD\ff = \xx_s \cdot \ff_s, \quad
  P\ff = (\ff \cdot \nn) \nn,
  \label{eqn:operators}
\end{align}
for the differential and projection operators. Then,
Equation~\eqref{eqn:vesVelocity} can be written as
\begin{align}
  \dot{\xx} &= \uu_\infty(\xx) + \beta P(\BB\xx + \TT\sigma)
  + \SS (\BB\xx + \TT\sigma), \quad \DD \dot{\xx} = 0.
\end{align}
We note that the operators defined in Equation~\eqref{eqn:operators}, as
well as the single-layer potential defined in Equation~\eqref{eqn:SLP}
all depend on the vesicle shape. A time stepping method that treats
these dependencies implicitly would result in a non-linear set of
equations that would be computationally expensive to solve. Instead, we
use a time stepping method that defines the operators at time $t^N$ and
discretizes the stiff terms due to high-order derivatives
implicitly~\cite{vee-gue-zor-bir2009, qua-bir2014}. In particular, the
first-order semi-implicit method that is applied iteratively with SDC is
\begin{alignat}{3}  
  \frac{\xx^{N+1} - \xx^N}{\triangle t} &= \uu_\infty(\xx^N) 
  + \beta P^N\left(\BB^N\xx^{N+1} + \TT^N\sigma^{N+1}\right) 
  + \SS^N\left(\BB^N\xx^{N+1} + \TT^N\sigma^{N+1}\right),  \\
  \DD^N\xx^{N+1} &= 1,
\end{alignat}
and this linear system is solved with matrix-free GMRES.

@@

\section{Hydrodynamics of A semipermeable inextensible membrane under strong confinement\label{sec:confinement}}
Here we consider a semipermeable inextensible membrane under strong
confinement: a long closely-fit channel and a contracting channel. In a
narrow microfluidic channel, a large pressure jump is often used to push
the vesicle through a closely-fit channel~\cite{abk-fai-sto2006}. Our
simulations show that such a configuration can lead to an amplification
of water permeation at a fast time scale of the order of
\todo[inline]{mini seconds?} mini seconds.  In addition, we find that a
large value of $\beta$ is needed to keep the membrane tension below the
threshold value for membrane poration. Such high permeability for water
may be regarded as an indication that the lipid bilayer membrane is
porated under strong confinement, and our model inextensible membrane is
for a porated vesicle within the Helfrich free energy framework. As in
the previous examples, we fix the membrane length and examine the
enclosed water content as a dynamic consequence of water flow, membrane
shape deformation, and membrane tension distribution.
